---
title: "2 - Vectors & Logical Expressions"
format: html
editor: visual
---

Welcome back to lesson two! In this lesson we will cover vectors (a one-dimensional data format), logical expressions and how to handle missing data.

# Vectors

In lession one, we learned about simple data types (numeric, character, ...). For example, if you measured the DBH of a tree, you may assign the numeric object to a variable DBH:

```{r}
dbh <- 8.42
```

However, often, you want to save multiple numerical objects in the same variable (e.g. the measured height of many different trees). To do this, you use the function `c()` to create a **vector**. The c stands for concatenate. For example:

```{r}
dbh_vector <- c(8.43, 16.22, 14.49, 19.28, 10.13, 7.23)
```

A vector holds **any number of ordered objects of the same data type**. You can use `str()` to see the data type and the length (number of objects) of a vector. You can also use `length()` to see the length of a vector.

```{r}
str(dbh_vector) # numeric vector with 6 elements
length(dbh_vector)
```

If the objects are not of the same type, R will automatically try to convert them to the same object (if this is not possible, there will be an error). Consider the following vector where all values are converted to the character data type:

```{r}
(my_vector <- c(10, "abcdefg", 9.2, TRUE))
str(my_vector)
```

::: callout-tip
If you add brackets around an assignment the object is automatically printed to the console!
:::

## Generating Vectors

You can generate vectors that contain ordered numbers (e.g. numbers 10 through 20) like this:

```{r}
(my_vector <- 1:20)
```

If you want to have a different step size, you can use the function `seq().`

```{r}
(my_vector <- seq(from = 1, to = 10, by = 0.5)) # vector from 1 to 10 with a step size of 0.5

(my_vector_2 <- seq(from = 5, to = 10, length.out = 7)) # vector with 7 equally spaced numbers between 5 and 10
```

Alternatively, you can use the function `rep()` to create a repeating vector.

```{r}
rep(7, times = 10) # repeat the number 7 ten times

# Note: You can give a vector as an argument to rep() to repeat the vector
rep(c(1, 2, 3), times = 10)

# Use "each" instead of "times" to order a vector as 1, 1, 1, 2, 2, 2... instead of 1, 2, 1, 2...
rep(c(1, 2, 3), each = 10)

# You can use rep() with non-numeric values
(detention <- rep("I must not tell lies.", times = 10))
```

## Vector Calculations

You can calculate with numeric vectors using the same arithmetic expressions from the first lesson. For example:

```{r}
# add 10 to every element of the vector
my_vector <- 1:5
my_vector + 10
```

You can also do calculations with two vectors of the same length.

```{r}
# Calculating volume of a cylinder
height <- c(108, 120, 138, 119)
radius <- c(14, 19, 28, 30)

# Note: pi is a built-in constant in R so we do not need
# to define it.
(volume <- pi * radius^2 * height)
```

If vectors have different lengths, the shorter vector will be *recycled*. This means it will be repeated to match the length of the longer vector; for example, in this case, `my_vector_2` is treated as `c(1, 2, 3, 1)`.

```{r, error = T}
my_vector_1 <- c(1, 2, 3, 4)
my_vector_2 <- c(1, 2, 3)

my_vector_1 - my_vector_2 # raises a warning, but calculation proceeds
```

## Selecting Vector Elements

To select an element of a vector, use square brackets with the number of the element you want to select.

```{r}
body_size <- c(10.4, 12.3, 8.1, 14.8)
body_size[2] # select the second element of body_size
```

You can select multiple elements by using a vector within the brackets:

```{r}
body_size[1:3]
body_size[c(2, 4)]
```

Or you can select all elements except specific ones:

```{r}
body_size[-1]
```

You can also replace the selected values.\

```{r}
body_size[1] <- 1000
body_size
```

## Logical Expressions & Operators

### Logical Expressions

A logical expression is an **expression that returns either TRUE or FALSE** (an object with the logical data type). Common logical expressions are the comparison expressions:

```{r}
4 > 10 # Greater than
10 >= 10 # Greater than or equal to
18 < 29 # Smaller than
12 < 7 # Smaller than or equal to
5 == 2 # Equals
5 != 3 # Does not equal
```

You can also use these operators with characters. Greater/Smaller Than then refers to alphabetical order.

```{r}
"cat" == "dog"
"cat" > "dog" # cat is earlier in the alphabet and therefore "smaller"
```

You can also use logical expressions with vectors. In these cases, the same rules as vector calculations apply:

```{r}
my_vector <- c(1, 2, 3, 4, 5)
my_vector > 3 # comparison to a single value 
my_vector == c(5, 4, 3, 2, 1) # comparison to another vector
```

Logical expressions with vectors always return a **logical vector** (vector with only TRUE/FALSE values). You can also use a TRUE/FALSE vector to select values from a vector:

```{r}
# select only value 1 and 5
selection_vector <- c(TRUE, FALSE, FALSE, FALSE, TRUE)
my_vector[selection_vector]
```

This is useful because it allows you to select values from a vector based on logical conditions. For example, if you are only interested in trees above a certain height:

```{r}
tree_heights <- c(10, 12, 7, 9, 11, 13)

# only select trees > 10 m height
# note that I have to write tree_heights twice
(tall_trees <- tree_heights[tree_heights >= 10])
```

You can also use the **%in% operator** with vectors. The %in% operator tells you whether the object on the left is in the object on the right:

```{r}
5 %in% c(1, 2, 3, 4, 5)

# with a vector on the left, each vector element
# is compared to the right
c(1, 2, 3, 4, 5) %in% 5

# you can also compare two vectors
c(3, 9, 2, 4, 8) %in% c(1, 2, 3, 4)
```

### Logical Operators

Logical operators **combine** multiple logical expressions. Logical operators recognized by R are AND (`&`), OR (`|`) and NOT (`!`).

```{r}
# AND
# For this expression to be TRUE,
# both sides need to be true
10 > 1 & 10 > 100

# OR
# For this expression to be TRUE,
# only one side needs to be TRUE
10 > 1 | 10 > 100

# NOT
# Not negates the value
# of the expression
!("cat" == "cat")

# same as:
"cat" != "cat"
```

You can use logical operators to select data with more criteria:

```{r}
tree_age <- c(2, 10, 9, 19, 21, 17, 4, 29)

# select only trees between age 10 and 20
tree_age[tree_age >= 10 & tree_age <= 20]
```

::: panel-tabset
## Comprehension Question

What values are selected in this expression?

`vector <- 1:20`

`vector[vector < 5 | vector > 15]`

## Answer

1, 2, 3, 4, 16, 17, 18, 19, 20
:::

## Functions and Vectors

You can use vectors as an argument in functions. You can use most functions you can use to manipulate single values (e.g. `round()` or `sqrt()`) with vectors as well.

```{r}
tree_dbh <- c(20.01, 18.73, 29.28, 14.93)
round(tree_dbh, digits = 1)
```

There are multiple functions that provide summary statistics of a vector:

```{r}
mean(tree_dbh) # mean
median(tree_dbh) # median
sd(tree_dbh) # standard deviation
max(tree_dbh) # maximum
which.max(tree_dbh) # which element is the maximum element?
min(tree_dbh) # minimum
range(dbh)

```

You can also use the function `sort()` to sort a vector:

```{r}
sort(tree_dbh)
sort(tree_dbh, decreasing = TRUE)
```

And you can use the function `any()` and `all()` with logical vectors to determine whether any/all values are TRUE.

```{r}
# Are there any DBHs over 20 cm?
any(tree_dbh > 20)

# Are all DBHs over 20 cm?
all(tree_dbh > 20)
```

## Missing Values / NA

Real data often contains missing values. For example, if you measured the temperature over the course of 5 days but had an instrument error on the 4th day, you may want to indicate this by setting the value as "missing". For this, you use `NA` ("Not Available").

```{r}
temperature <- c(10.3, 11.9, 9.3, NA, 12.1)
```

Using NA in calculations or logical expressions always results in NA.

```{r}
4 * NA
```

Many functions therefore don't work if the data includes missing values, but most of these functions take an optional argument that determines what to do with missing values. For example, the function `sum()` calculates the sum of all values in a vector and only works if the argument `na.rm` is set to TRUE.

```{r}
# does not work
sum(temperature)

# removes NAs before summing and therefore works
sum(temperature, na.rm = TRUE)
```

Since you can't use logical expressions to check for NAs, you can't use them to determine whether your data contains NA:

```{r}
NA == temperature
```

Instead, use the function `is.na()`. In combination with the `any()` function, you can see at a quick glance whether there are any missing values in your data.

```{r}
# see which values are NA
is.na(temperature)

# see if any values are NA
any(is.na(temperature))
```

::: {.callout-note collapse="true"}
## Advanced: Lists

NEEDS TO BE WRITTEN STILL
:::
