---
title: "3 - Data Frames & Factors"
format: html
editor: visual
---

Welcome to topic 3! In this lesson, we will cover data frames and matrices (2-dimensional data structures).

## Data Frames

You are already familiar with vectors (1-dimensional data structures). You may already have realized an issue with vectors: We can only save one variable per vector. If we measure multiple variables, for example the diameter at breast height (DBH), height, and species of a tree, we might instead want to record this information in a table with rows and columns. This is where **2-dimensional data structures** like data frames become relevant.

A data frame consists of rows and columns (similar to for example an excel table). Each column holds one vector: this means that every column can hold a different data type, but within a column the data type needs to be consistent. It is therefore good practice to have **one row per observation** (for example, one row per measured tree) and **one column per measured variable** (so, for example, one column for DBH, height and species respectively). The columns in a data frame are usually named; the rows are typically not (but can be named if necessary).

To create a data frame, use the `data.frame()` function.

```{r}
# creates a dataframe with 3 columns and 4 rows
trees <- data.frame(
  dbh = c(29, 16, 18, 21),
  height = c(10, 12, 23, 17),
  species = c("Beech", "Beech", "Fir", "Fir")
)

# you can use str() for a quick overview
# of a data frame
str(trees)
```

A data frame needs to have the same number of rows in every column:

```{r, error  = TRUE}

# Raises an error since the number of rows does not match
data.frame(
  var1 = c(1, 2, 3, 4),
  var2 = c("Blue", "Green", "Yellow")
)
```

::: callout-caution
If the number of rows of one column is a multiple of the number of rows of the other (for example, 2 and 4 or 30 and 90 rows), the shorter column will be recycled.
:::

## Selecting Data

To select data from a data frame, you can also use \[square brackets\]. However, since the data has two dimensions, you need to include both the row and the column number of the data you want to access. For this, remember **RoCo: Rows First, Columns Second.**

```{r}
trees[1, 3] # select the first row, third column

trees [2:3, c(1, 3)] # select row 2 and 3, column 1 and 3
```

If you want to select all rows/columns, you still need to add the comma, but not add a number.

```{r}
trees[, 3] # select all entries in the third column
```

If you columns have names, you can also **select columns by name.**

```{r}
trees[1, "height"] # first entry in the height column

trees[, c("height", "dbh")] # height & dbh column
```

If you only want to select a single column by name, you can use the \$ operator instead.

```{r}
trees$height # select the entire height column

trees$dbh[2] # select the second row of the dbh column
```

Just like with vectors, you can use logical expressions and operators to select data.

```{r}
# select trees with height > 20 m
trees[trees$height > 20, ]

# select trees with dbh < 20 and of species fir
trees[trees$dbh < 20 & trees$species == "Fir", ]
```

And you can replace selected values:

```{r}
trees[1, 3] <- "Pine"
```

## Summary Statistics

As an example for this lesson, we will use one of R's many **built-in datasets** (we will cover how to read in your own data instead soon). These datasets are useful for instruction and practicing your coding. To get an overview of all available data sets, run:

```{r}
library(help = "datasets")
```

In this case, we will use the `iris` data. This data contains measurements of 150 flowers of different iris species. For a more detailed description of the dataset, you can run `?iris`. To load the data, use:

```{r}
data(iris)
```

To gain a first overview of the data, you can use the `str()` command. Then use the `summary()` command for more detailed information, including the most common summary statistics (mean, maximum, minimum, quantiles).

```{r}
str(iris)

summary(iris)
```

## Factors

If we call `summary()` on our trees dataframe we can see that the information in the species column is not particularly helpful - we can see that there are four character objects, but not what they contain or how many trees of which species we have (like with the iris dataset above). This is because R does currently not recognize that these characters represent **categorical data.**

```{r}
summary(trees)
```

The categorical data type in R is called a **factor**. You can easily convert a variable into a factor using the `factor()` function.

```{r}
trees$species_factor <- factor(trees$species)

# Now the species_factor column provides a more useful summary
summary(trees)
```

Each factor has **levels** (categories). You can use the `levels()` function or `str()` to inspect them.

```{r}
levels(trees$species_factor)
str(trees)
```

Internally, a factor saves a number for each observation. The number represents the category; in this case, 1 represents "Beech" and 2 represents "Fir".

When converting a variable into a factor, you can optionally use the arguments `levels` and `labels`. By specifying "levels", you can change the order of your factor levels (otherwise, levels will be ordered alphabetically by default). With "labels", you can change the name of each factor level that will displayed in the summary statistics. Specifying levels and labels becomes especially relevant when plotting categorical data (covered in lesson 5).

For example, if you want to change the order of your level so that "Fir" is the first level and change the display names to their scientific names, you would use this code:

```{r}
# You can split up function arguments into multiple lines for readability
trees$species_factor <- factor(
  trees$species,
  levels = c("Fir", "Beech"),
  labels = c("Abies alba", "Fagus sylvatica")
  )
summary(trees)
```

## Renaming Columns

You can use the `names()` function to access the column names of a data frame.

```{r}
colnames(trees)
```

You can also use this function to rename the columns using an assignment:

```{r}
colnames(trees) <- c("dbh_cm", "height_m", "species_chr", "species_factor")
str(trees)
```
