---
title: "09 - Writing your own functions"
editor: visual
---

## Writing functions

You have already learned about functions in R - ranging in complexity from `sum()` to `ggplot()`. In addition to the many functions included in R and packages, you can write your own functions. This is useful if there is code you need to reuse often in different parts of your scripts.

To write your own functions, the syntax as follows:

``` r
my_function <- function(argument1, argument2){
  code of the function
  
  return(return_value)
}
```

The *return value* of a function is the new number created by a function - for example, the `sum()` function returns the sum of the values given to it. If you call `x <- my_function`, the return value will be assigned to x. However, functions are not required to have a return value. If you leave out the `return()` part, the function will simply return `NA`.

The following code creates a function `multiply_numbers()` that takes as arguments two numbers (a and b) and returns the two multiplied together.

```{r}
multiply_numbers <- function(a, b){
  result <- a * b
  return(result)
}
```

You can use your own functions as you would any other function in R.

```{r}
multiply_numbers(2, 3)
```

### Example: Calculating Species Diversity

Consider that you have conducted a survey where you counted tree species at five different sites.

```{r}
survey <- data.frame(
  site = c("A", "B", "C", "D", "E"),
  oak = c(12, 0, 8, 20, 5),
  maple = c(8, 15, 3, 0, 10),
  birch = c(5, 2, 12, 4, 9),
  pine = c(15, 18, 1, 0, 6)
)
str(survey)
```

Now you want to calculate a diversity index for each site to determine which one has the highest species diversity. We want to calculate Simpson's Diversity index, which is defined as:

$$D = \sum_{i=1}^{S}(n_i/N)^2$$

where `D` is the diversity index, `S` is the species richness, `N` is the number of individuals, and `n_i` is the number of individuals of a specific species. Already we can see that this is a mildly complex equation (no worries if you are unfamiliar with it - it isn't required prerequisite knowledge, just used here for illustrative purposes!). We therefore don't want to write it out for every site and should instead wrap it in a function.

```{r}
# function takes as arguments the species counts
calculate_simpson_index <- function(counts){
  species_richness = length(counts) # total number of species
  number_individuals = sum(counts) # how many individuals are there in total?
  proportions <- counts / number_individuals # calculate n_i / N
  
  # calculate the Simpson index
  index <- sum(proportions**2)
  
  # return value is the index (single value)
  return(index)
}
```

We can then easily use this to calculate the species richness of individual sites:

```{r}
# calculate species richness for site A
# we need to select column 2 - 5 since column 1 contains the site name
calculate_simpson_index(survey[1, 2:5])

# calculate species richness for site B
calculate_simpson_index(survey[2, 2:5])
```

We can also use our own function as the function argument in `apply()` (see chapter 8) and in other places that accept base R functions, such as `dplyr::summarize()`.

```{r}
# loop over rows (MARGIN = 1)
apply(survey[, 2:5], MARGIN = 1, calculate_simpson_index)
```

### Default Values

Some functions in R have default values. For example, the function `round()` has an argument `digits` that specifies how many digits the value is rounded to. Per default, digits is set to 0, but it can be overwritten.

```{r}
# default value
round(pi)

# overwriting default value
round(pi, digits = 2)
```

We can also give default values to our own functions. Consider the following function that returns `TRUE` if the given temperature is above the boiling temperature. By default, the boiling temperature is the boiling temperature of water (100 °C), but it can be changed so the function can be used for other substances.

```{r}
is_boiling <- function(temperature, threshold = 100){
  # is temperature equal to/greater than boiling threshold?
  boiling <- temperature >= threshold
  return(boiling)
}
```

```{r}
temperature_measurement <- 60

# default value
is_boiling(temperature_measurement)

# default value overwritten
is_boiling(temperature_measurement, threshold = 30)
```
